A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-15 natural history of scavenging.tex
added figure and cleaned up handling time

\begin{abstract} 
% Here, we chart the natural history of scavenging. 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. Chief among these are low-cost locomotion, high detection distances, effective carcass processing and a carrion-producing habitat. 
We use these four components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focussing on dominant vertebrate groups of the land, sea and air. 
The result is a document on the natural history of scavenging, the first to our knowledge. 
Our idea of a scale of scavenging can be applied to any species at any time to judge the importance of this behaviour in its diet. 
\end{abstract}

@@

\subsection*{The Difficulty of Scavenging}
The chief hurdle of scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. %TG: Two times "opportunities", replaced by "chances"
However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet their energetic demands.
%Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. %TG: make the following sentence more sense?
Once found, the scavenger must be able to out-compete any potential competitors and process the typically decaying carcass which contains microorganism derived toxins  \citep{ruxton2014fruit}.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics.
All of these facets are essentially the key parameters found in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. %TG: Jaysus, this sentence doesn't read well! "Functional response curves of optimal foraging theory" can probably be phrased in a nicer/easier way...
%With this framework we compare the ability of scavenging across vertebrates. % focusing on each of these facets and their relationship with scavenging.
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the importance of carrion in its diet.

@@

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
It seems that the use of soaring in extinct species as a means for scavenging is likely.

@@

We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. 
Similarly we would expect that extinct species would also scavenge in a similar fashion depending on the efficiency of their flight. 
For example, early birds such as \textit{Archaeopteryx} are predicted to have been poor, relatively inefficient fliers \citep{nudds2010narrow}. 
%Similarly something about the efficiency of pterosaurs flight. TG: Was this a comment? By the way, about pterosaurs (though it might not be necessary to mention it), most of them where tiny little feckers like Archaeopteryx or even smaller.
The importance of efficient flying over large areas may explain the lack of scavenging behaviour in bats as they are generally nocturnal, a time when they would receive no aid from convective air currents \citep{norberg2012vertebrate}. 
%smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), %AK Don't think it is the case that bats are less efficient than birds in terms of flight 
That said, \textit{Necromantis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) had a robust cranio-mandibular morphology, and is a likely candidate for scavenging behaviour \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.% amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}. %TG: we can cut this number of bat species to avoid getting into stamp collecting debate. + there are effectively a two shit load units of them which makes the Necromantis example not really convincing.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the likelihood of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Interestingly, style of swimmming in fish does not significantly affect the cost of movement \citep{williams1999evolution}. 
Though sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show large ranges (e.g. great white sharks (\textit{Carcharodon carcharias})) \citep{bruce2006movements} and it seems reasonable that they would use oceangraphic currents to further reduce movement costs  \citep{ruxton2004energetic}. %and are able to maintain a type of endothermy from thier movements \citep{bernal2015sharks}. %TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
Facultative scavenging is seen in many marine groups, including species of extant sharks like white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}.
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and Pliocene mysticete whales \citep[5.3 to 3.6 Mya;]{schwimmer1997scavenging,ehret2009caught}.

@@

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would especially prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya %TG: that early???
have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system. 
That said, aquatic endotherms have and do scavenge. 
For instance, early whales such as \textit{Basilosaurus} seem to fit into the same niche as killer whales \textit{Orcinus orca} and we have some evidence for scavenging in both \citep{fahlke2012bite,Whitehead415}.
%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic} as well as the relative inefficiency of gas exchange in mammals cf. birds and fish \citep{williams1999evolution}.
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often mentioned.
The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
Despite being intuitive, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Metabolic rate however, unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also be true of extinct species. 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

@@

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizeable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
We can (and have) use(d) these insights to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Million years ago; Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya) were likely to have been ambush predators which would argue against them taking a lot of carrion. 
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

@@

The best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

@@

Aside from sight, many birds have well developed olfactory systems \citep{AR:AR22815} including three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}).
% three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}.
An atuned sense of smell is obviously useful in detecting decaying carrion from the air.
% This would be impossible for the visually reliant old world species.

% Depending on the species, a carcass in water either floats or descends to the sea floor \citep{Whitehead415}.
In contrast to the air, aquatic species have to contend with the low-light environment where visual detection distances are far lower (< 100 m) than they would be in the air.
As such, aquatic animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderence of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed %TG: wind speed? Or current speed?
 determined the number of sharks feeding at whale carcasses, indicating they were dependent on detecting the odours from the decaying whales.

\section*{Handling Time}
Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over.
Being able to extract nutrients from remnants gives a scavenger a great advantage.
%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:
Thus, some phenotypic adaptations such as the ability to crush bone (as seen in hyenas) reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable %TG: is that an adjective?
meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}. %TG: I've removed the intelligence bit from the sentence.

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone e.g. the robust skull and dentition of \textit{Tyrannosaurus rex} \citep{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. %TG: A comment about the future cleaning up phase. I think mentions like the 570000 Newtons are not essential and could go to some supplementary files. Same applies for most of the physiques in the locomotion part, it is not essential in the narrative to have the number (yet important to link to them!).
This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}.

@@

The support of water allows for many aquatic species to reach large sizes thus granting the benefits. 
For example, white sharks, which grow up to 6 metres long, can be sustained by 30 kg of blubber for over six weeks \citep{carey1982temperature}.

>>>>

\begin{abstract} 
% Here, we chart the natural history of scavenging. 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. 
These can be thought of as some of the principal parameters in optimal foraging theory namely, encounter rate, handling time and prey availability. 
% Chief among these are low-cost locomotion, high detection distances, effective carcass processing and a carrion-producing habitat. 
We use these components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focussing on the dominant vertebrate groups of the land, sea and air. 
The result is a document on the natural history of scavenging, the first to our knowledge. 
Our idea of a scale of scavenging can be applied to any species at any time to judge the importance of this behaviour in its diet. 
\end{abstract}

@@

\subsection*{The Difficulty of Scavenging}
The chief hurdle to scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. %TG: Two times "opportunities", replaced by "chances"
However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet their energetic demands.
%Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. %TG: make the following sentence more sense?
Once found, the scavenger must be able to out-compete any potential competitors and process the, typically decaying, carcass replete with microorganism derived toxins  \citep{ruxton2014fruit}.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics.
All of these facets are essentially the key parameters found in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. %TG: Jaysus, this sentence doesn't read well! "Functional response curves of optimal foraging theory" can probably be phrased in a nicer/easier way...
%With this framework we compare the ability of scavenging across vertebrates. % focusing on each of these facets and their relationship with scavenging.
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the importance of carrion in its diet.

@@

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
It seems probable that extinct species used soaring as a means for scavenging.

@@

We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. 
Similarly we would expect that extinct species would also scavenge in a similar fashion depending on the efficiency of their flight. 
For example, early birds such as \textit{Archaeopteryx} are predicted to have been poor, relatively inefficient fliers \citep{nudds2010narrow} and so ill-suited to finding carrion. 
%Similarly something about the efficiency of pterosaurs flight. TG: Was this a comment? By the way, about pterosaurs (though it might not be necessary to mention it), most of them where tiny little feckers like Archaeopteryx or even smaller.
The importance of efficient flying over large areas may explain the lack of scavenging behaviour in bats as they are generally nocturnal, a time when they would receive no aid from convective air currents \citep{norberg2012vertebrate}. 
%smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), %AK Don't think it is the case that bats are less efficient than birds in terms of flight 
That said, \textit{Necromantis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) had a robust cranio-mandibular morphology, and is a likely candidate for scavenging behaviour \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.% amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}. %TG: we can cut this number of bat species to avoid getting into stamp collecting debate. + there are effectively a two shit load units of them which makes the Necromantis example not really convincing.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in (salmonid) fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the likelihood of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Interestingly, style of swimmming in fish does not significantly affect the cost of movement \citep{williams1999evolution}. 
Though sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show large ranges, e.g. great white sharks (\textit{Carcharodon carcharias}) \citep{bruce2006movements}, and it seems reasonable that they would use oceangraphic currents to further reduce movement costs  \citep{ruxton2004energetic}. %and are able to maintain a type of endothermy from thier movements \citep{bernal2015sharks}. %TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
Facultative scavenging is seen in many marine groups, including species of extant sharks like white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}. 
The former which grow up to 6 metres long, can be sustained by 30 kg of whale blubber for over six weeks \citep{carey1982temperature}. 
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and Pliocene (5.3 to 3.6 Mya) mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.

@@

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would especially prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya %TG: that early???
have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system. 
That said, aquatic endotherms have and do scavenge. 
For instance, early whales such as \textit{Basilosaurus} (38 to 36.5 Mya) seem to fit into the same niche as killer whales (\textit{Orcinus orca}) and we have some evidence for scavenging in both \citep{fahlke2012bite,Whitehead415}.
%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic} as well as the relative inefficiency of gas exchange in mammals cf. birds and fish \citep{williams1999evolution}.
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often mentioned.
The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
Despite being intuitive, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Metabolic rate however, unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also have been true of extinct species with the same physiology. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
%Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion. 
Although the earliest evidence of vertebrate scavenging comes from the Permian (298.9 - 252.17 Mya) where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}.

@@

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizeable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
These insights allow us to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Million years ago; Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya), were likely to have been ambush predators which would argue against them taking a lot of carrion. 
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

@@

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

@@

Aside from sight, many birds have well developed olfactory systems \citep{AR:AR22815} including three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}).
% three species of vultures within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}.
Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}.
An atuned sense of smell is obviously useful in detecting decaying carrion from the air over a heavily forested habitat.
% This would be impossible for the visually reliant old world species.

% Depending on the species, a carcass in water either floats or descends to the sea floor \citep{Whitehead415}.
In contrast to the air, aquatic species have to contend with a low-light environment where visual detection distances are far lower (< 100 m) than they would be in the air.
As such, aquatic animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderence of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed %TG: wind speed? Or current speed?
 determined the number of sharks feeding at whale carcasses, indicating they were dependent on detecting the odours from the decaying whales.

\section*{Handling Time}
Since the food a scavenger depends on is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be subject to decay as well as competition.
So being able to overcome competitors, maximse the nutrient gain from the remnants, and survive long enough between meals are all essential parts of carcass handling time. 
% Being able to extract nutrients from remnants gives a scavenger a great advantage.

%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:
In the ability to eat bone scavengers have arrived at a way to feed on a resource that is typcially too hard for many predators to process. 
% Thus, some phenotypic adaptations such as the ability to crush bone reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable %TG: is that an adjective?
meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth, enabling contemperaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}. %TG: I've removed the intelligence bit from the sentence.

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone e.g. the robust skull and dentition of \textit{Tyrannosaurus rex} \citep{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. %TG: A comment about the future cleaning up phase. I think mentions like the 570000 Newtons are not essential and could go to some supplementary files. Same applies for most of the physiques in the locomotion part, it is not essential in the narrative to have the number (yet important to link to them!).
\cite{rayfield2001cranial} note this as being powerful enough to break open skeletal material \citep{rayfield2001cranial}.

@@

The support of water allows for many aquatic species to reach large sizes thus granting its benefits. 
\cite{collins2005trends} found "contrasting relationships between size (body mass) and depth in the scavenging and predatory demersal ichthyofauna". 
Predatory species saw a reduction in body mass with depth whereas the reverse trend was true for scavengers. 
This, the authors pointed out, is because randomly distributed carrion is better exploited by fish with larger body sizes owing to starvation resistance.
